\section{PFLOTRAN Regression Test Manager}

The test manager for PFLOTRAN is a python program that is responsible
for reading a configuration file, identifying the tests declared in
the file, running PFLOTRAN on the appropriate input files, and then
comparing the results to a known ``gold standard'' output file.

\subsection{Running the Test Manager}
The test manager can be run in two ways, either as part of the build
system using ``make'' or manually.

There are two options for calling the test manager through make:
``make check'' and ``make test''. The ``check'' target runs a small
set of tests that verify that PFLOTRAN is built and running on a given
system. This would be run by user to verify that their installation of
PFLOTRAN is working. The ``test'' target runs a fuller set of
regression tests intended to identify when changes to the code cause
significant changes to PFLOTRAN's results.

Calling the test manager through make relies on make variables from
PETSc to determine the correct version of python to use, if PFLOTRAN
was build with MPI, and optional configurations such as unstructured
meshes. The version of python used to call the test manager can be
changed from the command line by specifying python:

\begin{verbatim}
$ cd ${PFLOTRAN_DEV}/src/pflotran
$ make PYTHON=/opt/local/bin/python3.3 check
\end{verbatim}

To call the test manager manually:
\begin{verbatim}
$ cd ${PFLOTRAN_DEV}/regression_tests
$ python regression-tests.py \
    --executable ../src/pflotran/pflotran \
    --config-file shortcourse/copper_leaching/cu_leaching.cfg \
    --tests cu_leaching
\end{verbatim}

Some important command line arguments when running manually are:
\begin{itemize}
\item executable : the path to the PFLOTRAN executable
\item mpiexec : the name of the executable for launching parallel
  jobs, (mpiexec, mpirun, aprun, etc).
\item config-file : the path to the configuration file containing the
  tests you want to run
\item recursive-search : the path to a directory. The test manager
  searches the directory and all its sub-directories for configuration
  files.
\item tests : a list of test names that should be run
\item suites : a list of test suites that should be run
\item update : indicate that the the gold standard test file for a
  given test should be updated to the current output.
\item new-tests : indicate that the test is new and current output
  should be used for gold standard test file.
\item check-performance : include the performance metrics ('SOLUTION'
  blocks) in regression checks.
\end{itemize}

The full list of command line options and a brief description can be
found by running with the ``--help'' flag:
\begin{verbatim}
$ python regression-tests.py --help
\end{verbatim}

\subsection{Test output}
The test manager produces terse screen output, only printing critical
warnings, a progress bar as each test is run, and a summary of the
overall test results. Example results from running \texttt{make test} are:

\begin{mdframed}
\small
\begin{verbatim}
  Test log file : pflotran-tests-2013-05-06_13-07-05.testlog

** WARNING ** : mpiexec was not provided on the command line.
                All parallel tests will be skipped!

Running pflotran regression tests :
..........S........SSSS.......F..........SSS
---------------------------------------------------------------------
Regression test summary:
    Total run time: 179.259 [s]
    Total tests : 44
    Skipped : 8
    Tests run : 36
    Failed : 1
\end{verbatim}
\normalsize
\end{mdframed}

The progress bar records a period, ``.'', for each successful test, an
``S'' if a test was skipped, an ``F'' for failed tests, a ``W'' for a
test that generated a warning, and an ``E'' for a test that generated
an internal error.

Each time the test suite is run, a log file is generated in the
regression test directory. The log file contains a detailed record of
every test run, including: the directory containing the test, the
command line call to PFLOTRAN used to run the test, a diff command to
compare the regression files, and a list of failures. The log file can
quickly be searched for ``skip'', ``fail'', ``error'' to identify the
tests that generated the message.

The test directories contain any files generated by PFLOTRAN during
the run. Screen output for each test is contained in the file
``\$\{TEST\_NAME\}.stdout''.



\subsection{Configuration Files}
The regression test manager reads tests specified in a series of
configuration files in standard ``cfg'' (or windows ``ini'' file)
format. They consist of a series of sections with key-value pairs:
\begin{verbatim}
[section-name]
key = value
\end{verbatim}
Section names should be all lower case, and spaces must be replaced by
a hyphen or underscore. Comments are specified by a ``\#'' character.

A test is declared as a section in the configuration file. It is assumed that
there will be a PFLOTRAN input file with the same name as the test
section. The key-value pairs in a test section define how the test is
run and the output is compared to the gold standard file. 
\begin{verbatim}
[calcite-kinetics]
#look for an input file named `calcite-kinetics.in'
np = 2
timeout = 30.0
concentration = 1.0e-10 absolute
\end{verbatim}

\begin{itemize}
\item np = N, (optional), indicates a parallel test run with N
  processors. Default is serial. If mpiexec in not provided on the
  command line, then parallel tests are skipped.
\item timeout = N, (optional), indicates that the test should be allowed to run
  for N seconds before it is killed. Default is 60.0 seconds.
\item TYPE = TOLERANCE COMPARISON, indicates that data in the
  regression file of type TYPE should be compared using a tolerance of
  TOLERANCE. Know data types are listed below.
\end{itemize}

The data types and default tolerances are:
\begin{itemize}
\item time = 5 percent
\item concentration = $1\times 10^{-12}$ absolute
\item generic = $1\times 10^{-12}$ absolute
\item discrete = 0 absolute
\item rate = $1\times 10^{-12}$ absolute
\item volume\_fraction = $1\times 10^{-12}$ absolute
\item pressure = $1\times 10^{-12}$ absolute
\item saturation = $1\times 10^{-12}$ absolute
\item residual = $1\times 10^{-12}$ absolute
\end{itemize}
The default tolerances are deliberately set very tight, and are
expected to be overridden on a per-test or per configuration file
basis.  There are three known comparisons: ``absolute'', for absolute
differences ($\delta\!=\!|c\!-\!g|$), ``relative'' for relative differences ($\delta\!=\!{|c\!-\!g|}/{g}$), and ``percent''
for specifying a percent difference ($\delta\!=\!100\!\cdot\!{|c\!-\!g|}/{g}$).

In addition there are two optional sections in configuration
files. The section ``default-test-criteria'' specifies the default
criteria to be used for all tests in the current file. Criteria
specified in a test section override these value. A section name
``suites`` defines aliases for a group of tests.
\begin{verbatim}
[suites]
serial = test-1 test-2 test-3
parallel = test-4 test-5 test-6
\end{verbatim}
Common test suites are \texttt{standard} and \texttt{standard\_parallel}, used by
``make test'', and domain specific test suites, \texttt{geochemistry},
\texttt{flow}, \texttt{transport}, \texttt{mesh}, et cetra.

\subsection{Creating New Tests}

We want running tests to become a habit for developers so that ``make
pflotran'' is always followed by ``make test.'' With that in mind, ideal
test cases are small and fast, and operate on a small subsection of
the code so it is easier to diagnose where a problem has
occurred. While it may (will) be necessary to create some platform
specific tests, we want as many tests as possible to be platform
independent and widely used. There is a real danger in having test
output become stale if it requires special access to a particular
piece of hardware, operating system or compiler to run.

The steps for creating new regression tests are:
\begin{itemize}
\item Create the PFLOTRAN input file, and get the simulation running
  correctly.

\item Tell PFLOTRAN to generate a regression file by adding a
  regression block to the input file, e.g.:
\begin{verbatim}
REGRESSION
  CELLS 
    1
    3978
  /
  CELLS_PER_PROCESS 4
END
\end{verbatim}

\item Add the test to the configuration file

\item Refine the tolerances so that they will be tight enough to
  identify problems, but loose enough that they do not create a lot of
  false positives and discourage users and developers from running the
  tests.

\item Add the test to the appropriate test suite.

\item Add the configuration file, input file and ``gold'' file to revision control.

\end{itemize}

\textbf{Guidelines for setting tolerances go here, once we figure out what to recommend.}

\subsection{Updating Test Results}

The output from PFLOTRAN should be fairly stable, and we consider the
current output to be ``correct''. Changes to regression output should
be rare, and primarily done for bug fixes. Updating the test results
is simply a matter of replacing the gold standard file with a new
file. This can be done with a simple rename in the file system:
\begin{verbatim}
mv test_1.regression test_1.regression.gold
\end{verbatim}
Or using the regression test manager:
\begin{verbatim}
$ python regression-tests.py --executable ../src/pflotran/pflotran \
    --config-file my_test.cfg --tests test_1 --update
\end{verbatim}
Updating through the regression test manager ensures that the output
is from your current executable rather than a stale file.

Please document why you updated gold standard files in
your revision control commit message.
